---
title: "p8105_hw2_yc3577"
author: "Yimeng Cai"
date: "10/01/2023"
output: github_document
---

```{r load libraries}
library(tidyverse)
library(dplyr)
library(readxl)
```

Problem 1

Step 1 clean & analyze pols-month data using read_csv to read the data, then use separate function to separate the mon variable into day year month. Then rename the month int data into month names. Finally use mutate to take values of gop and dem into president. Then select related columns and drop na.
```{r clean pols-month}
pols_df = 
  read_csv("fivethirtyeight_datasets/pols-month.csv") |>
  separate(mon, into = c('year', 'month', 'day'), sep = '-') 
month_names = 
  c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')
pols_df$month = month_names[as.integer(pols_df$month)]
pols_df = 
  mutate( pols_df,
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop"),
    year = as.integer(year)) |>
  select(-day, -starts_with("prez")) |>
  drop_na()
pols_df

```

Step 2 clean & analyze snp data using read_csv to read the data, then use separate function to separate the mon variable into day year month. Then formulate the variable type of month day and year. Use relocate to arrange the column in the orders of year and month. Then select related columns and drop na.
```{r Clean snp}
snp_df = 
  read_csv("fivethirtyeight_datasets/snp.csv") |>
  separate(date, into = c('month', 'day', 'year'), sep = '/') |>
  mutate(
    month = as.integer(month),
    day = as.integer(day),
    year = as.integer(year),
    month = month(month, label = TRUE),
    year = ifelse(year<16, year+2000, year+1900)) |>
  relocate(year, month) |>
  select(year, month, close) |>
  drop_na()

snp_df
```
Step 3 clean & analyze unemployment data using read_csv to read the data, then use separate function to separate the mon variable into day year month. Then formulate the variable type of month day and year. Use relocate to arrange the column in the orders of year and month. Then select related columns and drop na.
```{r Clean unemployment}
unem_df =  
  read_csv("fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "unemployment"
  )|>
  mutate(
    year = as.integer(year))|>
  select(year, month, unemployment) |>
  drop_na()

unem_df
```
```{r}
combined_df1 =
  left_join(pols_df, snp_df)|>
  left_join(x = _, y = unem_df)
combined_df1
```


Problem 2

Step 1 Clean and read Mytrash data using `read_excel` and `clean_names`. Then use `mutate` to calculate homes_powered by inputting the function `(weight_tons*500)/30`.  Use `add_column(type = "Mytrash")` to add additional variable to dataset for later combining usage. Finally drop NA data in dumpster with `drop_na(starts_with("dumpster"))`.
```{r}
Mrtrash_df = 
  read_excel("202309 Trash Wheel Collection Data.xlsx", sheet = 1, range = "A2:N587") |>
  janitor::clean_names() |>
  mutate(
    homes_powered = (weight_tons*500)/30) |>
  add_column(type = "Mrtrash")|>
  drop_na(starts_with("dumpster"))
Mrtrash_df
```

Step 2 Clean and read Proftrash data using `read_excel` and `clean_names`. Then use `mutate` to calculate homes_powered by inputting the function `(weight_tons*500)/30`. Then formulate the year column into character with `as.character(year)`. Use `add_column(type = "Proftrash")` to add additional variable to dataset for later combining usage. Finally drop NA data in dumpster with `drop_na(starts_with("dumpster"))`.
```{r}
Proftrash_df = 
  read_excel("202309 Trash Wheel Collection Data.xlsx", sheet = 2) |>
  janitor::clean_names() |>
  mutate(
    homes_powered = (weight_tons*500)/30,
    year = as.character(year))|>
  add_column(type = "Proftrash")|>
  drop_na(starts_with("dumpster"))

Proftrash_df
```

Step 3 Clean and read Gwytrash data using `read_excel` and `clean_names`. Then use `mutate` to calculate homes_powered by inputting the function `(weight_tons*500)/30`. Then formulate the year column into character with `as.character(year)`. Use `add_column(type = "Gwytrash")` to add additional variable to dataset for later combining usage. Finally drop NA data in dumpster with `drop_na(starts_with("dumpster"))`.
```{r}
Gwytrash_df = 
  read_excel("202309 Trash Wheel Collection Data.xlsx", sheet = 4) |>
  janitor::clean_names() |>
  mutate(
    homes_powered = (weight_tons*500)/30,
    year = as.character(year))|>
  add_column(type = "Gwytrash")|>
  drop_na(starts_with("dumpster"))

Gwytrash_df
```

Step 4 Combine `Mytrash_df`, `Proftrash_df` using 1st `bind_rows`. Then combine binded dataframe with `Gwytrash_df` using 2nd `bind_rows`.
```{r}
combine_df2 = 
  bind_rows(Mrtrash_df, Proftrash_df) |>
  bind_rows(x = _, y = Gwytrash_df)
combine_df2
```


Problem 3

Step 1 Read and Clean the `MCI_baseline.csv` of baseline data using `read_csv` and then use `janitor::clean_names()` to clean the variable names. During table reading, we need to skip the 1st row with `skip = 1`. Then we need to mutate the variable of sex and apoe4 with `mutate(sex = recode( sex = recode(sex, "0" = "F", "1" = "M"), apoe4 = recode(apoe4, "0" = "apoe4_noncarrier", "1" = "apoe4_carrier")` to make sure sex and apoe4 are appropriately encoded into character format, and change the '.' into 'NA' by `age_at_onset = ifelse(age_at_onset == ".", NA, age_at_onset)` for later fileter steps. Finally, we need to use filter of `filter(current_age < age_at_onset | is.na(age_at_onset))` to remove any participants who do not meet the stated inclusion criteria. 
```{r}
base_df = 
  read_csv("/Users/christinecym/Desktop/R data/P8105/HW/P8105_hw2_yc3577/data_mci/MCI_baseline.csv", skip =1) |>
  janitor::clean_names()|>
  mutate(
    sex = recode(sex, "0" = "F", "1" = "M"),
    apoe4 = recode(apoe4, "0" = "apoe4_noncarrier", "1" = "apoe4_carrier"),
    age_at_onset = ifelse(age_at_onset == ".", NA, age_at_onset)
    )|>
  filter(current_age < age_at_onset | is.na(age_at_onset))
  

base_df
```

Step 2 Read and Clean the `mci_amyloid.csv` of biomarkers using `read_csv` and then use `janitor::clean_names()` to clean the variable names. During table reading, we need to skip the 1st row with `skip = 1`. Then we need to rename the variable name with `rename(id = ..., time_0 = ...)` to make sure we have a common name id, and same format of time_X by change baseline to the time_0. Finally, we need to use `pivot_longer(time0:time8, names_to = ..., values_to = ...)` to make a tidy table with a column named `"period_time"` that includes all the time_period into one column as a new variable. Then finally drop na data with `drop_na()`. 
```{r}
amy_df = 
  read_csv("/Users/christinecym/Desktop/R data/P8105/HW/P8105_hw2_yc3577/data_mci/mci_amyloid.csv", skip =1) |>
  janitor::clean_names() |>
  rename(
    id = study_id,
    time_0 = baseline
  )|>
  pivot_longer(
    time_0:time_8,
    names_to = "period_time",
    values_to = "year") |>
  drop_na()

amy_df
```

Step 3 Combine two tables use `anti_join(base_df, amy_df)` based on the common ID, since we need to find the only participants in each datasets. We need to use anti_join twice to find each differences. (`anti_join(amy_df, base_df)`). Then we need to use `inner_join(base_df, amy_df)` base on the common ID to find the participants in both datasets. 
```{r}
combine_df3 =
  anti_join(base_df, amy_df)
combine_df3

combine_df4 = 
  anti_join(amy_df, base_df)
combine_df4

combine_df5 =
  inner_join(base_df, amy_df)
combine_df5
```

